---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

To review all the cyrrilic text: File -> Reopen with Encoding...(UTF-8)

```{r}
df <- read.csv("electricity.csv", stringsAsFactors = T)
head(df)
```

```{r}
ggplot(data = df)
```
```{r}
# ggplot2 is built on graphics grammar
# aes = aestetics = atributes of geometries
# до естетик в першу чергу належить положення, в aes х - це положення

ggplot(df, aes(x = consumption))
```
```{r}
# now we add geometries
ggplot(df, aes(x = consumption))+
  geom_histogram()

```


```{r}
# other way to do the same - aes are placed in geometry
ggplot(df) + geom_histogram(aes(x = consumption))
  
```

```{r}
# a cse when we want to have 2 geometries
ggplot(df) + geom_histogram(aes(x = consumption)) + geom_rug(aes(x = consumption))
# OR
ggplot(df, aes(x = consumption)) + geom_histogram() + geom_rug()
```
As we have outliers, we need to exclude them.

```{r}
df <- df[df$consumption > 0,]
ggplot(df, aes(x=consumption)) + 
  geom_histogram(binwidth = 300, fill= "darkcyan")
```

```{r}
# here we change the scale
# scale_x_continuous(breaks = seq(10000, 24000, 2000)) = де буде розташований графік
# limits= c(10000, 20000) = limits of datarange
ggplot(df, aes(x=consumption)) + 
  geom_histogram(binwidth = 500, fill= "darkcyan") +
  scale_x_continuous(breaks = seq(10000, 24000, 2000), limits= c(10000, 20000)) +
  scale_y_continuous(breaks = seq(0, 900, 100))
```

```{r}
# we can add categorical data as fill
# position of how the categories are places in in geom_histogram
ggplot(df, aes(x=consumption, fill = season)) + 
  geom_histogram(binwidth = 500) +
  scale_x_continuous(breaks = seq(10000, 24000, 2000)) +
  scale_y_continuous(breaks = seq(0, 900, 100))
```
```{r}
# adding facet through facet_wrap(~season)
ggplot(df, aes(x=consumption, fill = season)) + 
  geom_histogram(binwidth = 500, position = ) +
  scale_x_continuous(breaks = seq(10000, 24000, 2000)) +
  scale_y_continuous(breaks = seq(0, 900, 100)) +
  facet_wrap(~season)
```
```{r}
# we an edo faced from pairs of categorical veriables

ggplot(df, aes(x=consumption, fill = season)) + 
  geom_histogram(binwidth = 500, position = ) +
  scale_x_continuous(breaks = seq(10000, 24000, 2000)) +
  scale_y_continuous(breaks = seq(0, 900, 100)) +
  facet_wrap(month~season)
```
```{r}
# working with 2 data sets
ggplot(df, aes(x=consumption)) + 
  geom_histogram(data = df[, 1:9],
                 binwidth = 500, fill= 'gray', alpha = 0.8) +
  geom_histogram(binwidth = 500, fill='darkcyan') +
  scale_x_continuous(breaks = seq(10000, 24000, 2000)) +
  scale_y_continuous(breaks = seq(0, 900, 100)) +
  facet_wrap(~season)+
  theme_dark()
```

```{r}
ggplot(df, aes(x=consumption, fill = season)) + 
  geom_histogram(binwidth = 500) +
  scale_x_continuous(breaks = seq(10000, 24000, 2000)) +
  scale_y_continuous(breaks = seq(0, 900, 100)) +
  scale_fill_brewer(type = 'seq') 
```


```{r}
ggplot(df, aes(x = as.Date(datetime), y = consumption))+geom_line()

ggplot(df, aes(x = as.Date(datetime), y = consumption))+
  geom_point(color='darkcyan', alpha = 0.5)

ggplot(df, aes(x = as.Date(datetime), y = consumption))+
  geom_point(color='darkcyan', alpha = 0.5, position = 'jitter')

ggplot(df, aes(x = as.Date(datetime), y = consumption, color = consumption))+
  geom_point( alpha = 0.5, position = 'jitter') 

ggplot(df, aes(x = as.Date(datetime), y = consumption, color = hour))+
  geom_point( alpha = 0.5, position = 'jitter') +
  scale_color_distiller(type = 'div')
```

```{r}
ggplot(df, aes(x = hour, y = consumption)) +
  geom_point(alpha = 0.8)

ggplot(df, aes(x = factor(hour, ordered = T, levels= c(0:23)), y = consumption)) +
  geom_boxplot()
```

```{r}
# making datetime a valid datetime format(was factor or string)
df$datetime <- as.POSIXct(df$datetime)

# making beutiful graph
ggplot(df)+
  geom_tile(aes(x = as.Date(df$datetime), 
                y = hour, fill = consumption)) +
  scale_x_date(name = 'date', date_breaks = '1 month', date_labels = '%m') +
  scale_fill_distiller(direction = -1, type ='div')+
  theme_minimal()
```

