---
title: "Data exploration"
subtitle: Hanna Pylieva
output:
  pdf_document: default
  html_notebook: default
---

Loading earthquakes data for year 2017.
```{r}
dat <- read.csv("earthquakes.csv", stringsAsFactors = T)
head(dat)
```

At first we need to understand the dataset.
Let's look at the columns we have and their types.
The description of columns is available here: https://earthquake.usgs.gov/earthquakes/feed/v1.0/csv.php 
```{r}
str(dat)
```
Find the absolute number of NAs in columns.
```{r}
colSums(is.na(dat))
```
And the fractions in % of NAs for each column are:
```{r}
colSums(is.na(dat)/nrow(dat)*100)
```


Now let's look at the data...
General distribution of eathquakes on the Earth during the last year. 

```{r warning=FALSE}
library(dplyr)
library(ggplot2)
library(maps)
source("utils.R")
```

```{r}
by_place <- dat %>% group_by(longitude, latitude, place) %>% summarize(count = n()) %>% arrange(count)
```

```{r}
mdat <- map_data('world')

ggplot() + 
  geom_polygon(dat=mdat, aes(long, lat, group=group)) +
  geom_point(data=by_place, aes(x=longitude, y=latitude , size = count, col = count) , alpha = 0.2) + 
  scale_size(range = c(0, 10)) +
  ggtitle("Count of EQs in 2017") +
  scale_color_gradient(low = 'gold1', high = 'red1') +
  theme_dark() +
  theme(plot.title = element_text(hjust = 0.5)) 

```

### Let's understand how our records are distributed by categorical variables.  
```{r}
colnames <- as.data.frame(colnames(dat))
```

```{r}
# alternatively
# ggplot(dat, aes(status)) + geom_bar(aes(y = (..count..)/sum(..count..))) + ggtitle("Status") + labs(y= "percentage")+
  # theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))

p1 <- ggplot(dat, aes(status)) + geom_bar() + ggtitle("Count EQs by Status ") +
  geom_text(stat='count', aes(label=..count..),position = position_dodge(width = 1), vjust = -0.5, size = 3) +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))

by_status_perc <- dat %>% group_by(status) %>% summarise(percentage = n() / nrow(dat))
p2 <- ggplot(by_status_perc, aes(x = status, y = percentage, label=sprintf("%0.2f%%", round(percentage*100, digits = 2)))) +   
    geom_bar(stat = "identity") + ggtitle("Ratio EQs by Status") + 
    geom_text(position = position_dodge(width = 1), vjust = -0.5, size = 3) +
    theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
multiplot(p1, p2, cols = 2)


by_type<- dat %>% group_by(type) %>% summarize(count = n(), percentage = n() / nrow(dat))
by_type$type <- factor(by_type$type, levels = by_type$type[order(-by_type$count)] )

ggplot(by_type, aes(x = type, y = count, label=sprintf("%0.2f%%", round(percentage*100, digits = 2)))) + 
  geom_bar(stat = "identity" ) + ggtitle("EQ type") + 
  geom_text(aes(y = count) ,position = position_dodge(width = 1), vjust = -0.5, size = 3) +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))


by_magtype <- dat %>% group_by(magType) %>% summarize(count = n(), percentage = n() / nrow(dat))
by_magtype$magType <- factor(by_magtype$magType, levels = by_magtype$magType[order(-by_magtype$count)] )
ggplot(by_magtype, aes(x = magType, y = count, label=sprintf("%0.0f%%", round(percentage*100, digits = 2)))) + 
  geom_bar(stat = "identity") +ggtitle("Magnitude Type") + 
  geom_text(aes(y = count) ,position = position_dodge(width = 1), vjust = -0.5, size = 3.5) +
  theme(plot.title = element_text(hjust = 0.5))

by_magSource <- dat %>% group_by(magSource) %>% summarize(count = n(), percentage = n() / nrow(dat))
by_magSource$magSource <- factor(by_magSource$magSource, levels = by_magSource$magSource[order(-by_magSource$count)] )
ggplot(by_magSource, aes(x = magSource, y = count, label=sprintf("%0.0f%%", round(percentage*100, digits = 2)))) + 
  geom_bar(stat = "identity") +ggtitle("Magnitude Source") + 
  geom_text(aes(y = count) ,position = position_dodge(width = 1), vjust = -0.5, size = 3.5) +
  theme(plot.title = element_text(hjust = 0.5))

by_locationSource <- dat %>% group_by(locationSource) %>% summarize(count = n(), percentage = n() / nrow(dat))
by_locationSource$locationSource <- factor(by_locationSource$locationSource, 
                                           levels = by_locationSource$locationSource[order(-by_locationSource$count)] )
ggplot(by_locationSource, aes(x = locationSource, y = count, label=sprintf("%0.0f%%", round(percentage*100, digits = 2)))) + 
  geom_bar(stat = "identity") +ggtitle("Location Source") + 
  geom_text(aes(y = count) ,position = position_dodge(width = 1), vjust = -0.5, size = 3.5) +
  theme(plot.title = element_text(hjust = 0.5))

by_net <- dat %>% group_by(net) %>% summarize(count = n(), percentage = n() / nrow(dat))
by_net$net <- factor(by_net$net, levels = by_net$net[order(-by_net$count)] )
ggplot(by_net, aes(x = net, y = count, label=sprintf("%0.0f%%", round(percentage*100, digits = 2)))) + 
  geom_bar(stat = "identity") + ggtitle("Net") + 
  geom_text(aes(y = count) ,position = position_dodge(width = 1), vjust = -0.5, size = 3.5) +
  theme(plot.title = element_text(hjust = 0.5))

```
Summary:  
1. Most of EQs (94%) were reviewded by people.  
2. Most of events (97.5%) are eqrthquakes, all the rest types of events have ration less than 1% in the total number of events.  
3. 63% of events have local(ml) magnitude type (predumably with magnitude range 2 - 7.5).  
25% of events were of magnitude type Duration (magnitude range <4) and 9% of events were of type Short-period surface wave (magnitude range 3.5-7)  
4. It looks like it is the same network which originally authored both magnitude and location of the event (as distributions of magSource/locationSource/net by networks are similar). The networks which authored the most of events (magSource/locationSource/net) is ak (31% of events). It is interesting, events of which location did each network authored.  

```{r}
# hjust - horizontal justification
ggplot() + 
  geom_polygon(dat=mdat, aes(long, lat, group=group)) +
  geom_point(data=dat, aes(x=longitude, y=latitude ,  col = net), alpha = 0.5, size = 0.5) +
  ggtitle("Events by networks which authored them") +
  # theme_dark() +
    scale_fill_distiller(palette = "Spectral")+
  theme(plot.title = element_text(hjust = 0.5)) 
```
Looks like ak network is a local one for Alaska region, whereas us networ reports about event worldwide.  
Let's look at the picture without us net.
```{r}
dat_no_us <- dat %>% filter(net != 'us')
ggplot() + 
  geom_polygon(dat=mdat, aes(long, lat, group=group)) +
  geom_point(data=dat_no_us, aes(x=longitude, y=latitude ,  col = net), alpha = 0.5, size = 0.5) +
  ggtitle("Events by networks which authored them", subtitle = "'us' net excluded") +
  theme(plot.title = element_text(hjust = 0.5))
```

So **us** is the only network which reports about events worldwide. All the rest of networs are local.  It's also interesting that  that us network has reported only 14% of events, but as they are very distributed on the map one can think from the first glance that us network reported 80% of events. This also tells us that there are areas which need special attention (like Alaska), and there local networks were established. And the number of events on the rest of world is relatively smaller.  


### Let's not look at float features and their distribution.
```{r}
breaks = seq(-12,10,by=2) 
ggplot(dat, aes(x=mag)) + geom_histogram(breaks = breaks) +
  stat_bin(geom="text", aes(label=sprintf("%0.2f%%",round(..count../sum(..count..)*100,2))),
           position = position_dodge(width = 1), vjust = -0.5,
           breaks = breaks) +
  scale_x_continuous(breaks = breaks) +
  ggtitle("Distribution of magnitude of events") +
  theme(plot.title = element_text(hjust = 0.5),panel.grid.minor = element_blank())
```
Let's look how trustfull are measurements of magnitude in general.  
```{r}
breaks = seq(0,2,by=0.05)
ggplot(dat, aes(x=magError)) + geom_histogram(breaks=breaks) +
  stat_bin(geom="text", aes(label=sprintf("%0.0f%%",round(..count../sum(..count..)*100,2))),
           position = position_dodge(width = 1), vjust = -0.5,
           breaks = breaks, size = 2.5) +
  ggtitle("Distribution of Magnitude Error", subtitle = "The percentage on labels is taken among rows with magError non NA") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), panel.grid.minor = element_blank())
```
Pay attention that 41280 rows (32.5% of dataset) contain NA values and are not depicted on the plot above. The NA values mean that the  contributing seismic network does not supply uncertainty estimates.

```{r}
breaks = seq(0,200,by=10) 
ggplot(dat, aes(x=depth)) + geom_histogram(breaks = breaks) +
  stat_bin(geom="text", aes(label=sprintf("%0.0f%%",round(..count../sum(..count..)*100,2))),
           position = position_dodge(width = 1), vjust = -0.5,
           breaks = breaks, size = 3) +
  scale_x_continuous(breaks = breaks) +
  ggtitle("Distribution of depth of events") +
  theme(plot.title = element_text(hjust = 0.5), panel.grid.major.x = element_blank(), panel.grid.minor = element_blank())
```
Let's underestand how trustfull are measurements of depth in general.
```{r}
breaks = seq(0,35,by=1) 

ggplot(dat, aes(x=depthError)) + geom_histogram(breaks = breaks) +
  stat_bin(geom="text", aes(label=sprintf("%0.0f%%",round(..count../sum(..count..)*100,2))),
           position = position_dodge(width = 1), vjust = -0.5,
           breaks = breaks, size = 3) +
  scale_x_continuous(breaks = seq(0,35,by=2) ) +
  ggtitle("Distribution of depthError") +
  theme(plot.title = element_text(hjust = 0.5),panel.grid.minor = element_blank())
```
Most depth is identified as less than 10 (almost 60% of data), whereas from data description we can see that the depth could have been set to a fixed value and 5 or 10 km are often used in mid-continental areas and on mid-ocean ridges. But anyway we can see that most of the EQs are shallow (depth less than 60km).   
The error in depth up to 1km mostly makes depth trustful enough from my point of view.

### Now I want to look at data in dynamics

```{r}
dat$time <- as.POSIXct(dat$time)
by_day <- dat %>% group_by(date = as.Date(time)) %>% summarise(count = n())

ggplot(by_day, aes(x = date, y = count)) +  geom_line() +
  ggtitle("Number of events in dynamics during the year 2017") +
  scale_x_date(name = 'date', date_breaks = '1 month', date_labels = '%b %y') +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(hjust = 0.5))
```
So during the year 2017 the highest number of EQs observed in total on the Earth was on the first half of May. The  relatively small number was on Feb-Mar, second half of May, beginning of June, Dec.  
It is interesting to look at distribution of EQs on earth om monthly basis. But my computer wasn't able to do that on facets:) That's why let's look at least how the number in distributed by networks who fixed the event.
```{r}
by_day_net<- dat %>% group_by(date = as.Date(time), net) %>% summarise(count = n()) 
ggplot() + 
  #geom_polygon(dat=mdat, aes(long, lat, group=group), fill="grey50") +
  geom_line(data=by_day_net, aes(x=date, y = count, col = net)) +
  ggtitle("Events by networks which authored them") +
  theme(plot.title = element_text(hjust = 0.5))

by_day_net<- by_day_net %>% filter(net %in% c('ak','us','nc'))
ggplot() + 
  #geom_polygon(dat=mdat, aes(long, lat, group=group), fill="grey50") +
  geom_line(data=by_day_net, aes(x=date, y = count, col = net)) +
  ggtitle("Events by networks which authored them") +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(hjust = 0.5))

```

So now we can see that that extreme number of EQs in the first half of may was on Alaska region. The growth of number of EQs in July-August is due to intensive EQs reported by **nc** network which corresponds to Nothern California Seismic System.  

Just curious how the magnitude changed on the events fixed by **ak** station (will filter by magSOurce as we are interested exactly in magnitude distribution for now).

```{r}
ak_magSource <- dat %>% filter(magSource == 'ak') %>%
          mutate(mony= paste(format(time, "%m"), year = format(time, "%Y"), sep = "-")) 

ggplot(ak_magSource, aes(x = mony, y = mag)) + geom_violin() +
  #geom_point(binaxis='y', stackdir='center', dotsize=0.1, col="darkblue", alpha = 0.5) +
   geom_jitter( alpha = 0.01, col="darkblue", width = 0.5) +
   ggtitle("Distribution of events' magnitude for events reported by Alaska Earthquake Center", subtitle = "Year 2017") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))

```
As we can see the magnitude fixed by **ak** network is more or les similarly distributed on all months across the year 2017 except for May 2017, which has the highest median. So there were not only a lot of EQs on May 2017, but they also were of a magnitude slighly higher than average throughout the year. 